---
title: "Analysis"
author: "Chi Lam Poon"
date: "4/18/2021"
output: html_document
---

### Loading data

The DESeq2 object was already created previously.

```{r load}
suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(ggplot2)
  library(edgeR)
  library(ggbeeswarm)
  library(ggpubr)
})
suppressMessages(dds <- readRDS("/athena/angsd/scratch/zhp4001/ANGSD-project/data/deseq.obj.rds"))
```

### DESeq!

```{r design1}
dds@design
dds <- DESeq(dds)
anno <- metadata(dds)[["annotation"]]
# try to get DEGs for TOD
res.tod <- results(dds)
res.tod
# take a look
resOrdered.tod <- res.tod[order(res.tod$padj),] %>% data.frame()
resOrdered.tod$geneName <- anno[match(rownames(resOrdered.tod), anno$id),]$symbol
head(resOrdered.tod)
DESeq2::plotMA(res.tod, ylim=c(-2,2))
```

```{r design2}
# try to get DEGs for sex
res.sex <- results(dds, contrast = c("gender","male","female"))
res.sex
resOrdered.sex <- res.sex[order(res.sex$padj),] %>% data.frame()
resOrdered.sex$geneName <- anno[match(rownames(resOrdered.sex), anno$id),]$symbol
head(resOrdered.sex)

DESeq2::plotMA(res.sex, ylim=c(-5,5))
```

Now I try to use data with no Y genes because :

```{r noY}
dds2 <- readRDS("/athena/angsd/scratch/zhp4001/ANGSD-project/data/deseq.obj.noXY.rds")
dds2@design
dds2 <- DESeq(dds2)

# try to get DEGs for TOD
res.tod2 <- results(dds2)
res.tod2
# take a look
resOrdered.tod2 <- res.tod2[order(res.tod2$padj),] %>% data.frame()
resOrdered.tod2$geneName <- anno[match(rownames(resOrdered.tod2), anno$id),]$symbol
head(resOrdered.tod2)

# try to get DEGs for sex
res.sex2 <- results(dds2, contrast = c("gender","male","female"))
res.sex2
resOrdered.sex2 <- res.sex2[order(res.sex2$padj),] %>% data.frame()
resOrdered.sex2$geneName <- anno[match(rownames(resOrdered.sex2), anno$id),]$symbol
head(resOrdered.sex2)

DESeq2::plotMA(res.sex2, ylim=c(-5,5))
```

Decided to continue the analysis using data with XY genes included...

```{r shrinkage}
library(apeglm)
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="gender_male_vs_female", type="apeglm")
resLFC
DESeq2::plotMA(resLFC, ylim=c(-5,5))
```

#### Volcano plot
```{r vol}
resOrdered.sex$color <- ifelse(abs(resOrdered.sex$log2FoldChange)>=log2(2) & resOrdered.sex$padj <= 0.05, "steelblue", "lightgrey")

ggplot(resOrdered.sex, aes(x=log2FoldChange, y=-log10(padj), color=color)) +
  geom_point(size=1.5) + theme_bw() + scale_color_identity() + 
  geom_vline(xintercept=c(-log2(2), log2(2)), alpha=0.6, linetype="dotted") +
  geom_hline(yintercept=-log10(0.05), alpha=0.8, color="steelblue", linetype="dotted") +
  ggtitle("Male vs Female") +
  annotate("text", x=-10, y=230, label="XIST", fontface = 'italic') +
  annotate("text", x=12, y=110, label="DDX3Y", fontface = 'italic') +
  annotate("text", x=16, y=80, label="USP9Y", fontface = 'italic') +
  annotate("text", x=-3, y=40, label="TSIX", fontface = 'italic')
```

## Pathway analysis
I just threw the significant DEGs (log2FC >= |log2(1.5)| & adjusted.Pval <= 0.05) into gProfiler.
```{r path}
pos.ids <- rownames(resOrdered.sex[resOrdered.sex$color=="steelblue" & resOrdered.sex$log2FoldChange>0, ])
pos.ids <- gsub(".[.*$]", "", pos.ids)
neg.ids <- rownames(resOrdered.sex[resOrdered.sex$color=="steelblue" & resOrdered.sex$log2FoldChange<0, ])
neg.ids <- gsub(".[.*$]", "", neg.ids)
write.table(pos.ids, "/athena/angsd/scratch/zhp4001/ANGSD-project/data/sig.pos.id.txt", quote = F, row.names = F, col.names = F)
write.table(neg.ids, "/athena/angsd/scratch/zhp4001/ANGSD-project/data/sig.neg.id.txt", quote = F, row.names = F, col.names = F)

```

## Swarm plots

```{r splot}
cpms <- cpm(assay(dds))
log.cpms <- log2(cpms) + 1

plotting <- function(mat, gene, name) {
  meta <- colData(dds) %>% data.frame()
  gene.cpm <- log.cpms[rownames(log.cpms)==gene,]
  gene.cpm[gene.cpm==-Inf] <- 0
  df <- data.frame(obj=names(gene.cpm), cpm=gene.cpm, gender=meta$gender)
  
  ggplot(df, aes(gender, cpm, color=gender)) + geom_quasirandom() + theme_bw() +
    ggtitle(name) + ylab("log2(cpm)")
}

p1 <- plotting(log.cpms, "ENSG00000229807.12", "XIST")
p2 <- plotting(log.cpms, "ENSG00000270641.1", "TSIX")
p3 <- plotting(log.cpms, "ENSG00000067048.17", "DDX3Y")
p4 <- plotting(log.cpms, "ENSG00000114374.13", "USP9Y")
ggarrange(p1,p2,p3,p4,ncol=2, nrow=2,legend = "right", common.legend = T)
```

## Biotypes

```{r biotype}
resOrdered.sex$biotype <- anno[match(rownames(resOrdered.sex), anno$id),]$biotype
types <- resOrdered.sex[resOrdered.sex$color=="steelblue",]$biotype
table(types)
```

## Chromosomes for genes
```{r chrom}
resOrdered.sex$chrom <- anno[match(rownames(resOrdered.sex), anno$id),]$chrom
chroms <- resOrdered.sex[resOrdered.sex$color=="steelblue",]$chrom
tmp <- table(chroms)
sum(tmp[!names(tmp) %in% c("chrX","chrY")])
chr.df <- data.frame(chr=c("Autosomal", "X", "Y"), count=c(1648,41,27))
ggplot(chr.df, aes(x=chr,y=count,fill=chr)) +
  geom_bar(position="stack", stat="identity") + theme_bw()
```
